---
title: "Onset, regression"
author: "Marcel de Leeuw, (c) GeneCreek 2020"
date: "4/4/2020"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = F)

library(tidyverse)
library(ggpubr)
library(pROC)
library(mctest)
```

# aGvHD onset combination dataset
We combined data sets SRP243841, ERP105598 and SRP234378 into a 68 patients and 80 samples dataset "ON", comprising samples taken at aGvDH onset or beyond with matched controls. 

```{r}
load("../data/ON_mt.rda")
load("../data/ON_relaimp_status.rda")
```

# Biological safety level
At one month after stem cell infusion, the average biological safety level is higher in aGvHD cases than in controls.

```{r fig.width=2.5, fig.height=2, fig.align='center'}

comparisons <- list(c("no GvHD", "GvHD"))
ggplot(ON_mt, aes(x=status, y=bsl)) +
  geom_boxplot(alpha = 0.50, outlier.shape = NA) +
  geom_jitter(width = 0.2, shape=21, 
              aes(colour=status=="no GvHD", fill=status=="no GvHD")) +
  stat_compare_means(comparisons = comparisons, size=3) +
  ylim(1, 2.1) +
  theme_classic() + ylab("BSL") +
  theme(axis.line=element_blank(), legend.position="none",
        text=element_text(family="Open Sans", size=10))

```

# Linear regression
But we are also interested in individual species which could be used as biomarkers, provided we can "explain" a substantial part of the aGvHD/control variability through a combination of a limited number of species. To this end we use the R package relaimpo, which we have incorporated in an in-house algorithm, returning a rich data object. With the settings used, we tested 117 species.

```{r}
ON_relaimp_status$tested
```

The final model contains 20 independent variables.

```{r}
ON_relaimp_status$model
```

The total (unadjusted) R-squared exceeds 0.5...

```{r}
ON_relaimp_status$R2
```

...which breaks down as follows into individual variable contributions...

```{r}
ON_relaimp_status$regress
```

...which can be displayed as a bar chart:

```{r}
ON_relaimp_status$regress %>%
  ggplot(aes(x=regressor, y=R2, fill=R2<0)) +
  geom_bar(stat="identity") +
  coord_flip() + xlab(NULL) +
  theme_classic() + 
  annotate("text", x=7.5, y=-.075, parse=T, 
           label=sprintf("R^2: %.3f", ON_relaimp_status$R2)) +
  theme(axis.line=element_blank(),
        legend.position="none",
        text=element_text(family="Open Sans", size=10))

```

We can apply this linear model to the dataset, and get a ROC analysis with the following:

```{r}
coeffs <- ON_relaimp_status$model$coefficients[-1]
names(coeffs) <- gsub("\\.", " ", names(coeffs))
response <- coeffs %*% as.matrix(ON_species_comb[names(coeffs),ON_relaimp_status$data$sampleID])
ON_mt_response <- ON_relaimp_status$data %>%
  dplyr::mutate(resp=as.numeric(response)+
                  ON_relaimp_status$model$coefficients[1])

ggplot(ON_mt_response, aes(x=resp, fill=status!="GvHD")) + 
  geom_histogram(binwidth=.1)

ON_roc <- pROC::roc(ON_mt_response$status, ON_mt_response$resp)
pROC::ggroc(ON_roc) +
  geom_segment(aes(x=1, xend=0, y=0, yend=1), color="grey", linetype="dashed") + 
  annotate("text", x=.75, y=.55, label=sprintf("AUC: %.3f", ON_roc$auc))

```

